Cataclysmic variables are a key population of gravitational wave sources for LISA

### 2023-05-03 src/tex/ms.tex
update from overleaf and change figure color schemes

% Don't change these lines
\begin{document}
\label{firstpage}
\pagerange{\pageref{firstpage}--\pageref{lastpage}}
\maketitle

@@

The binary stars \lisa\ is sensitive to are those with orbital periods ranging from minutes to hours. One of the most important and best-studied populations in this period range are the so-called AM~CVn stars. These are ultra-compact binaries in which a white dwarf (WD) accretes material from a fully or partially degenerate companion (e.g. another WD). AM~CVn stars have orbital periods in the range $5~\mathrm{min} \lesssim P_{orb} \lesssim 65~\mathrm{min}$, near the "sweet-spot" of \lisa's sensitivity curve. Several AM~CVn systems have been identified as so-called \lisa\ verification binaries (\citealt{kupfer18,kupfer23}) as their gravitational wave amplitude, and consequently characteristic strain, are strong enough to be resolvable by \lisa\ within the first 4 years of operation. AM~CVn stars are also part of a larger Galactic population of ultra-compact binaries that includes, for example, detached double-degenerate systems. This population is expected to give rise to a broad-band, unresolved GW signal between $\approx 10^{-4}$ -- $10^{-2}$ Hz, which will be detectable by \lisa\, but will also limit the mission's sensitivity to other GW sources in this band \citep{nelemans01, ruiter10, nissanke12, korol17, lamberts19, breivik20, korol22}.

@@

There orbital period distribution of CVs contains two distinctive features, as illustrated in Figure~\ref{fig:porb}. First, the there is an obvious dearth of CVs in the "period gap" between $2~\mathrm{hr} \lesssim P_{orb} \lesssim 3~\mathrm{hr}$. Second, there is a "period spike" near the minimum period of $P_{min} \simeq 80~\mathrm{min}$. Both of these features are associated with changes in the structure of the donor star. When a CV reaches the upper edge of the period gap, its donor becomes fully convective. Such stars cannot support the roughly dipolar magnetic fields found in partially radiative stars, since those fields are anchored in the tachocline (the interface between the radiative and convective zone). As a result, the magnetic field topology of the donor must change at this point, and it is thought that MB-driven AML is also severely reduced. This causes a temporary loss of contact, allowing the donor to shrink back to its MS radius. The system then evolves through the period gap as a detached binary until the AML due to GR (and perhaps also due to residual MB or other mechanisms) has shrunk the orbit sufficiently for contact to be re-established. Mass transfer then resumes at the lower edge of the gap.

@@

Given their (relatively) high space density and the distinctive nature of their period distribution, CVs merit a close look as potential GW sources for \lisa\. We are aware of only two earlier attempts to do this. First, \cite{HBW90} included CVs as one of the six types of Galactic binary systems for which they estimated GW signals. They found them to be a significant, but sub-dominant population compared to W~UMa stars and "close WD binaries" (they did not explicitly consider semi-detached -- i.e. AM~CVn systems -- as part of the latter group). Second, \cite{MAA00} explicitly considered CVs as a population of interest for \lisa\. They predicted the GW signals for systems in a catalogue of CVs available at the time, finding that some of these ought to be detectable by the mission.

@@

Since CVs orbit with $f_{\rm{GW}} < 1\,\rm{mHz}$, their orbital evolution occurs on long-enough timescales that LISA is unlikely to observe any changes to the GW frequency of any given source. This means that the total signal to noise ratio can be successfully approximated as
\begin{equation}\label{eq:SNR}
    SNR = \frac{ASD}{\sqrt{S_n}} = \frac{\mathcal{A}\sqrt{T_{\rm{obs}}}}{\sqrt{S_n}},
\end{equation}
\noindent where $T_{\rm{obs}}$ is the observation duration and $S_n$ is the power spectral density sensitivity of \lisa. We use LEGWORK to compute these values (\citealt{LEGWORK_apjs,LEGWORK_joss}).

@@

\cite{pala20} presents the first volume-limited sample of CVs selected using parallax measurements provided by \gaia. The sample consists of $42$ CVs within $150\,\rm{pc}$, and is quantified to be $\approx 75\%$ complete. The sample is used to investigate the intrinsic properties of the Galactic CV population, and more specifically to infer the space density of CVs. Taking into account the missing systems due to completeness, the space density is found to be $\rho_0\simeq4.8 \times 10^{-6}$pc$^{-3}$. This value is based on the assumption that the CV population follows an axisymmetric Galactic disc profile of the form
\begin{equation}
\label{eq:disk}
    \rho = \rho_{0} \exp \left(-\frac{|z|}{h}\right)
\end{equation}

\noindent where the CV population scale height $h$ is taken to be 280pc and $z$ is the distance above the Galactic plane. In reality the scale height $h$ may increase for older CV systems (those with smaller orbital periods). However, the local CV population does not allow inference of $h(P_{\rm{orb}})$ due to relatively low numbers of observed CVs. Nevertheless, \cite{pala20} quote different space density values for different scale height assumptions, which are all found to be within a factor of 2. Here we employ the quoted canonical space density measurement of $\rho_0=4.8 \times 10^{-6}$pc$^{-3}$ and a scale height of $h=280$pc.

In order to calculate the GW signature of the nearby CV population, we use an empirically motivated scheme to assign orbital periods based on the \citet{pala20} $150\,\rm{pc}$ sample, then assuming a fiducial accretor mass of $0.75\,\rm{M_{\odot}}$, assign a donor mass based on the evolutionary track from \citet{knigge11}. As the orbital period of \gaia\ J154008.28$-$392917.6 is the only unknown in the sample, we arbitrarily set this to 81 minutes.

We augment this set by generating the missing systems from the \cite{pala20} sample. These simulated systems are distributed to follow the axisymmetric Galactic disk profile and measured space density from \cite{pala20}. To assign orbital periods to the simulated systems we attempt to resample the known orbital period distribution from observed CV systems in the \cite{pala20} sample. We first smooth this distribution to account for the observed low number of systems using a kernel density estimator. We then sample \textbf{12} systems from this smoothed orbital period distribution. Figure \ref{fig:porb} shows the orbital period distribution of the \cite{pala20} sample, as well as the 150pc sample used here (containing \textbf{54} systems, of which 42 are shared with\cite{pala20}). Similarly to the observed sample, we assign the donor mass based on the revised tracks of \cite{knigge11} assuming $m_1=0.75$M$_{\odot}$.

@@

The results of computing the characteristic strain after 4 years of \lisa\ operations are shown in Fig. \ref{fig:asd}, together with \textbf{THE SENSITIVITY CURVE}. Although several systems are found to lie close to the conventionally adopted \lisa\ sensitivity limit, 3 systems in particular attain a signal-to-noise detection $>4\sigma$. Unsurprisingly, these are the 3 closest systems, namely WZ Sge, VW Hyi and EX Hya. It is also interesting to note that there may be the possibility of detecting objects through their GW emission which have been missed due to electromagnetic detection biases and not included in the 150pc volume-limited sample.

@@

Understanding the Galactic population of binaries that \lisa\ will be sensitive to is crucial for correctly modelling and quantifying the \lisa\ sensitivity.

Note that both 3 individually resolvable systems, as well as the amplitude of the noise excess are expected to be lower estimates due to potentially under-represented population of period bouncers that are not included here.

Note assumption of constant scale height in this work. Possibility of resolving space density as a function of scale height and Porb?

@@

% Don't change these lines
\bsp	% typesetting comment
\label{lastpage}
\end{document}

>>>>

% Don't change these lines
\begin{document}
% \label{firstpage}
% \pagerange{\pageref{firstpage}--\pageref{lastpage}}
\pagerange{?--?}
\maketitle

@@

The binary stars \lisa\ is sensitive to are those with orbital periods ranging from minutes to hours. One of the most important and best-studied populations in this period range are the so-called AM~CVn stars. These are ultra-compact binaries in which a white dwarf (WD) accretes material from a fully or partially degenerate companion (e.g. another WD). AM~CVn stars have orbital periods in the range $5~\mathrm{min} \lesssim P_{orb} \lesssim 65~\mathrm{min}$, near the "sweet-spot" of \lisa's sensitivity curve. Several AM~CVn systems have been identified as so-called \lisa\ verification binaries (\citealt{kupfer18,kupfer23}) as their gravitational wave amplitude, and consequently characteristic strain, are strong enough to be resolvable by \lisa\ within the first four years of operation. AM~CVn stars are also part of a larger Galactic population of ultra-compact binaries that includes, for example, detached double-degenerate systems. This population is expected to give rise to a broad-band, unresolved GW signal between $\approx 10^{-4}$ -- $10^{-2}$ Hz, which will be detectable by \lisa\, but will also limit the mission's sensitivity to other GW sources in this band \citep{nelemans01, ruiter10, nissanke12, korol17, lamberts19, breivik20, korol22}.

@@

The orbital period distribution of CVs contains two distinctive features, as illustrated in Figure~\ref{fig:porb}. First, the there is an obvious dearth of CVs in the "period gap" between $2~\mathrm{hr} \lesssim P_{orb} \lesssim 3~\mathrm{hr}$. Second, there is a "period spike" near the minimum period of $P_{min} \simeq 80~\mathrm{min}$. Both of these features are associated with changes in the structure of the donor star. When a CV reaches the upper edge of the period gap at $\approx 3$ hr, its donor is thought to become fully convective. Such stars cannot support the roughly dipolar magnetic fields found in partially radiative stars, since those fields are anchored in the tachocline (the interface between the radiative and convective zone). As a result, the magnetic field topology of the donor must change at this point, and it is thought that MB-driven AML is also severely reduced. This causes a temporary loss of contact, allowing the donor to shrink back to its MS radius. The system then evolves through the period gap as a detached binary until the AML due to GR (and perhaps also due to residual MB or other mechanisms) has shrunk the orbit sufficiently for contact to be re-established. Mass transfer then resumes at the lower edge of the gap.

@@

Given their (relatively) high space density and the distinctive nature of their period distribution, CVs merit a close look as potential GW sources for \lisa. We are aware of only two earlier attempts to do this. First, \cite{HBW90} included CVs as one of the six types of Galactic binary systems for which they estimated GW signals. They found them to be a significant, but sub-dominant population compared to W~UMa stars and "close WD binaries" (they did not explicitly consider semi-detached -- i.e. AM~CVn systems -- as part of the latter group). Second, \cite{MAA00} explicitly considered CVs as a population of interest for \lisa. They predicted the GW signals for systems in a catalogue of CVs available at the time, finding that some of these ought to be detectable by the mission.

@@

Since CVs orbit with $f_{\rm{GW}} < 1\,\rm{mHz}$, their orbital evolution occurs on long-enough timescales that \lisa\ is unlikely to observe any changes to the GW frequency of any given source. This means that the total signal to noise ratio can be successfully approximated as
\begin{equation}\label{eq:SNR}
    SNR = \frac{ASD}{\sqrt{S_n}} = \frac{\mathcal{A}\sqrt{T_{\rm{obs}}}}{\sqrt{S_n}},
\end{equation}
\noindent where $T_{\rm{obs}}$ is the observation duration and $S_n$ is the power spectral density sensitivity of \lisa. We use LEGWORK to compute these values (\citealt{LEGWORK_apjs,LEGWORK_joss}).

@@

\cite{pala20} presents the first volume-limited sample of CVs selected using parallax measurements provided by \gaia. The sample consists of $42$ CVs within $150\,\rm{pc}$ and is estimated by them to be $\approx 75\%$ complete. Figure~\ref{fig:porb} shows the orbital period distribution for these 42 systems in both cumulative and differential forms.

Below, we will want to generate mock CV samples whose orbital periods are consistent with that of the intrinsic CV population, but are not restricted to the set of 42 measured values. In addition, we want the mock orbital periods to respect the period minimum near $P_{min} \simeq 80$~min and the period gap between $2~\mathrm{hr} \lesssim P_{orb} \lesssim 3~\mathrm{hr}$ \citep[e.g.][]{gansicke09, knigge06}. Both of these well-known features are present in the \cite{pala20} 150-pc sample, but are not as cleanly defined as in larger (but incomplete) samples. We have therefore approximated the observed cumulative distribution with a simple analytical function:
\begin{equation}
CDF(x) = 
\begin{cases}
      0 & P_{orb} < 76.78~\mathrm{min}, \\
      1.98 \left(\frac{x_1 + x_1^{3/4}}{1 + x_1 + x_1^2}\right) & 76.78 \mathrm{min} \leq P_{orb} \leq 2.10~\mathrm{hr},\\
      0.834 & 2.10~\mathrm{hr} < P_{orb} < 3.09~\mathrm{hr},\\
      0.26 \left(\frac{x_2 + x_2^{3/4}}{1 + x_2 + x_2^2}\right) + 0.834  & 3.09~\mathrm{hr} \leq P_{orb} \leq 10.58~\mathrm{hr},\\
      1 & P_{orb} > 10.58~\mathrm{hr}
\end{cases}
\label{eq:cdf}
\end{equation}
where $x = \log_{10}({P_{orb,m}})$, $x_1 = x - 1.885$, $x_2 = x - 2.269$
and $P_{orb,m}$ is the orbital period expressed in minutes. This cumulative  distribution is illustrated by the blue curve in the top panel of Figure~\ref{fig:porb}. We can then generate mock orbital periods from this via inversion sampling, i.e. by generating $R \sim U(0,1)$, a uniformly distributed random number between zero and unity, and then choosing $P_{orb}$ such that $CDF(x) = R$. The blue curve in the bottom panel of Figure~\ref{fig:porb} shows the resulting differential period distribution.

\cite{pala20} have used their 150-pc sample to investigate the intrinsic properties of the Galactic CV population, and, more specifically, to estimate the space density of CVs. Taking into account the missing systems due to completeness, the space density is found to be $\rho_0\simeq4.8 \times 10^{-6}$pc$^{-3}$. This estimate is based on the assumption that the CV population follows an axisymmetric Galactic disc profile of the form
\begin{equation}
\label{eq:disk}
    \rho = \rho_{0} \exp \left(-\frac{|z|}{h}\right)
\end{equation}

\noindent where the CV population scale height $h$ is taken to be 280~pc and $z$ is the distance above the Galactic plane. In reality the scale height $h$ may increase for older CV systems (those with smaller orbital periods). However, the local CV population does not allow inference of $h(P_{\rm{orb}})$ due to relatively low numbers of observed CVs. Nevertheless, \cite{pala20} quote different space density values for different scale height assumptions, which are all found to be within a factor of 2. Here we employ the quoted canonical space density measurement of $\rho_0=4.8 \times 10^{-6}$~pc$^{-3}$ and a scale height of $h=280$~pc.

In order to calculate the GW signature of the nearby CV population, we first consider the 42 systems in the \citet{pala20} 150-pc sample. All but one of these have known orbital periods. The only exception is \gaia\ J154008.28$-$392917.6, which is thought to be a  WZ~Sge-like system. We therefore set its orbital period to 81 minutes, close to $P_{min}$. For simplicity, we adopt an accretor mass of $m_1 = 0.75\ ~ \rm{M_{\odot}}$ for all systems and assign the corresponding donor mass $m_2$ based on the revised evolutionary track from \citet{knigge11}.

We then correct for the incompleteness of the \citet{pala20} 150-pc sample by simulating systems that are distributed to follow the axisymmetric Galactic disk profile and measured space density of $\rho_0=4.8 \times 10^{-6}$pc$^{-3}$ and scale height of $h=280$pc from \cite{pala20}. Our realisation of this mock population yields an additional \textbf{8} systems within 150-pc. The orbital periods of these additional CVs are drawn from the period distribution described by Equation~\ref{eq:cdf}. As for the observed systems, we adopt $m_1=0.75$~M$_{\odot}$ for all mock CVs and assign the donor mass based on the revised tracks of \cite{knigge11}.

@@

The results of computing the characteristic strain after four years of \lisa\ operations are shown in Fig. \ref{fig:asd}, together with the corresponding \lisa\ sensitivity curve within the frequency range of interest (\textbf{REF?}). Although several systems are found to lie close to the conventionally adopted \lisa\ sensitivity limit, 3 systems in particular attain a signal-to-noise detection $>4\sigma$. Unsurprisingly, these are the 3 closest systems, namely WZ Sge, VW Hyi and EX Hya. It is also interesting to note that there may be the possibility of detecting other CV systems through their GW emission which have been missed due to electromagnetic detection biases and not included in the current 150pc volume-limited sample.

@@

GW astronomy can be considered in its infancy, yet the prospects and opportunities this will bring to our understanding of both our Galaxy and the Universe will be transformational. Our observations and related interpretation of the GW sky will be heavily reliant on our understanding of both background and foreground unresolved GW signals. This is particularly important in resolving GW signals in the lower signal-to-noise regime. In this Letter we have shown how CVs will contribute to the expected signals to be detected by upcoming space-based GW observatories such as \lisa\ and \textit{TianQin} (\citealt{luo16}), and possibly to the \textit{Lunar Gravitational Wave Antenna} (\citealt{harms21}).

In this context, we have shown how some of the closest CVs should be individually resolvable within the first four years of \lisa\ operations. We further demonstrate how the Galactic CV population will significantly contribute to the \lisa\ mission sensitivity. In this respect we predict a characteristic GW noise excess which will increase in GW strain and rise above the currently simulated Galactic populations in the $\approx 0.2$-$0.4$ mHz GW range. This noise excess is directly related to the relatively high CV space density and CV evolutionary timescale properties.

As CVs evolve towards lower orbital periods they "pile-up" at the so called orbital period minimum. The consequence of this are a very characteristic GW noise excess that sharply diminishes at frequencies above $\approx 0.4$ mHz. 
In this work we have simulated the GW signals based on the observed electromagnetic population of CVs. The amplitude of the observed noise excess by e.g. \lisa\ can be used to directly constrain the CV space density, without the intrinsic biases affecting electromagnetic derived CV populations. For example, the so-called "period gap" where the donor stars in CVs temporarily detach and mass transfer onto the WD is quenched within the orbital period range $2~\mathrm{hr} \lesssim P_{orb} \lesssim 3~\mathrm{hr}$, is thought to be a purely electromagnetic bias to our observations. The prospect of studying the CV population with \lisa\ is also exciting in testing this conventional hypothesis.

@@

% Don't change these lines
\bsp	% typesetting comment
% \label{lastpage}
\end{document}
